White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2016-07-14 papers/wdmerger_II/paper.tex

%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
We consider the collisions of white dwarfs as possible progenitors of Type Ia 
supernovae.

\end{abstract}
\keywords{supernovae: general - white dwarfs}

@@

Over the past few years a number of groups have performed simulations of WD collisions
to understand whether they may yield astrophysical transients that look like SNe Ia
\citep{rosswog:2009,raskin:2010,loren-aguilar:2010,hawley:2012,garcia-senz:2013,
kushnir:2013,papish:2015,holcomb:2015}. Head-on collisions rapidly convert a significant
amount of kinetic energy into thermal energy in a small region and thus set up the
conditions for a thermonuclear detonation, and these previous simulations have indeed
found that detonations occur and convert a large amount of carbon/oxygen material into
iron-group elements.

We consider the collisions of white dwarfs as possible progenitors of Type Ia 
supernovae, using the code \castro\ \cite{castro}.

@@

Next is \aproxthirteen\ \citep{timmes:1999,timmes:2000}. This includes
all of the isotopes of \isoseven, and all of the $alpha$-chain particles between
silicon and nickel (${}^{32}$S, ${}^{36}$Ar, ${}^{40}$Ca, ${}^{44}$Ti, ${}^{48}$Cr,
and ${}^{52}$Fe). This network was used by \citet{hawley:2012} and \citet{raskin:2010}.
\citet{loren-aguilar:2010} and \citet{garcia-senz:2013} used a very similar network
that included additionally ${}^{60}$Zn. The \aproxnineteen\ network \citep{timmes:1999}
builds on \aproxthirteen\ by including isotopes for hydrogen burning and explicit
tracking of photodisintegration into ${}^{54}$Fe. This network was used by
\citet{kushnir:2013}, \citet{kushnir:2014}, \citet{rosswog:2009} for their
calculations with FLASH, and \citet{papish:2015}. Finally we will also
consider \aproxtwentyone, which includes all of the above plus ${}^{56}$Cr
and ${}^{56}$Fe and related reaction links. The primary virtue of using
the latter two networks is that they allow us to track changes away from
$Y_e = 0.5$.

@@

>>>>

%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
Collisions of white dwarfs are a subset of white dwarf mergers that occur
when the stars approach each other nearly head-on. These have the potential
for explosive nucleosynthesis, leading to a thermonuclear detonation that
disrupts both stars, possibly causing a Type Ia supernova. Simulation of
these events is challenging due to the rapid evolution of the nuclear burning,
posing numerical challenges related to both spatial and temporal resolution.
In this study we explore a number of the algorithms that affect the outcome
of reacting hydrodynamics simulations that lead to explosive burning and
address the question of what needs to be done to ensure that these simulations
are properly converged. We find that achieving a converged simulation in the
code \castro\ is non-trivial and requires demanding computational resources.
We also discuss the quantities obtained from our simulations that bear relevance
to observations of these events, particularly production of nickel-group
elements and gravitational wave emission during the collision.
\end{abstract}
\keywords{supernovae: general - white dwarfs}

@@

Over the past few years a number of groups have performed simulations of WD collisions
to understand whether they may yield astrophysical transients that look like SNe Ia
\citep{rosswog:2009,raskin:2010,loren-aguilar:2010,hawley:2012,garcia-senz:2013,
kushnir:2013,papish:2015,holcomb:2015}. Head-on collisions rapidly convert a significant
amount of kinetic energy into thermal energy in a small region and thus set up the
conditions for a thermonuclear detonation, and these previous simulations have indeed
found that detonations occur and convert a large amount of carbon/oxygen material into
iron-group elements. These papers varied significantly in the hydrodynamic methods used
(Lagrangian versus Eulerian methods), the methodology used for coupling nuclear
reactions to the hydrodynamics (including variation in the number of isotopes in
the nuclear network and in the evolution of the temperature), and the temporal and
spatial resolution. Consequently they yielded estimates of production of nickel-group
elements that varied significantly, leaving much uncertainty about how such an event
would appear observationally and whether it bears any resemblance to a SN Ia.

In this paper we seek to add another view to the problem using the reactive hydrodynamics
code \castro\ \citep{castro} and its underlying adaptive mesh refinement library \boxlib.
This paper is a follow-up to \citet{wdmergerI} (hereafter, Paper I), which established
the hydrodynamic methdology we are using to simulate mergers of white dwarfs.
Paper I left out a discussion of the methodology we are using for nuclear reactions,
and we fill out this information below in \autoref{sec:numericalmethods}. While it is
our aim is to add information to the discussion about collisions of white dwarfs,
we also view this work as a methodology study that asks to what extent we
can trust the reliability of explosive thermonuclear reactions in Eulerian
hydrodynamics simulations. As a corollary, to what extent does static and/or
adaptive mesh refinement resolve the types of problems that occur in low resolution
simulations? We discuss this in Section \autoref{sec:resolution}, though first
we examine the response of the code to a number of user-specified parameters in
\autoref{sec:parameters}. We then close with our thoughts on what our work implies
for the broader astrophysical reacting hydrodynamics literature in \autoref{sec:conclusion}.

@@

Next is \aproxthirteen\ \citep{timmes:1999,timmes:2000}. This includes
all of the isotopes of \isoseven, and all of the $\alpha$-chain particles between
silicon and nickel (${}^{32}$S, ${}^{36}$Ar, ${}^{40}$Ca, ${}^{44}$Ti, ${}^{48}$Cr,
and ${}^{52}$Fe). This network was used by \citet{hawley:2012} and \citet{raskin:2010}.
\citet{loren-aguilar:2010} and \citet{garcia-senz:2013} used a very similar network
that included additionally ${}^{60}$Zn. The \aproxnineteen\ network \citep{timmes:1999}
builds on \aproxthirteen\ by including isotopes for hydrogen burning and explicit
tracking of photodisintegration into ${}^{54}$Fe. This network was used by
\citet{kushnir:2013}, \citet{kushnir:2014}, \citet{rosswog:2009} for their
calculations with FLASH, and \citet{papish:2015}. Finally we will also
consider \aproxtwentyone, which includes all of the above plus ${}^{56}$Cr
and ${}^{56}$Fe and related reaction links. The primary virtue of using
the latter two networks is that they allow us to track changes away from
$Y_e = 0.5$.

@@

Paper I established our methodology for calculating the gravitational wave signature
of a white dwarf merger using the post-Newtonian framework of \citet{blanchet:1990}.
While a typical merger event has a characteristic profile including a chirp and ringdown,
white dwarf collisions feature sharply different gravitational wave signatures, and
hence may be interesting targets for observation for future low frequency gravitational
wave detectors \citep{eLISA}. The main caveat for these events compared to events
such as neutron star mergers is that due to the event being inherently less energetic,
the typical distance scale on which we could observe a white dwarf merger or collision
is the size of our own galaxy, with extragalactic events being unlikely to be
detected, at least with the first generation of space-based gravitational wave
detectors. \autoref{fig:gw_signal} is a plot of the gravitational wave strain for
the representative case of \autoref{sec:parameters}. We use the convention that the
absolute scale of the strain is normalized to a distance of 10 kpc (the strain scales
inversely with the source distance). The shape of the profile agrees qualitatively
with the gravitational wave signature predicted by \cite{loren-aguilar:2009:collisions},
though because they used an unequal mass pair the results do not match exactly.

\begin{figure}
  \centering
  \includegraphics[scale=0.425]{plots/gw_signal}
  \caption{Gravitational wave strain for the head-on collision of two $0.64\ \msolar$
           white dwarfs. The strain is normalized to a source distance of 10 kpc.
           \label{fig:gw_signal}}
\end{figure}

%TODO: use a 3D collision for this?
